---
title: "Simulated LMER"
author: "Dan Feeney"
date: "8/19/2021"
output:
  html_document: default
  pdf_document: default
---

## setting up enviornment 
```{r}
# https://www.tjmahr.com/plotting-partial-pooling-in-mixed-effects-models/ provided a lot o
# basic code and inspiration for this. While journals won't allow citing a blog, this was a great
# resource
rm(list=ls())
library(tidyverse)
library(lme4)
library(emmeans)
library(MuMIn)
library(patchwork)
set.seed(5280)

```


# Simulating data
```{r}
# set fixed effect parameters. https://journals.sagepub.com/doi/full/10.1177/2515245920965119
# set seed
set.seed(5280)

# set fixed effect parameters. https://journals.sagepub.com/doi/full/10.1177/2515245920965119

beta_0 <- 800 # intercept; i.e., the grand mean. this would be avg pwr

beta_1 <- 50 # slope; the effect of category this would be difference in pwr

# set random effect parameters

tau_0 <- 50 # by-subject random intercept sd. We assume this comes from a normal dist with  mean 0 and unknown SD. SD for random intercept by sub

omega_0 <- 1 # by-item random intercept sd for seated and standing. by condition random intercept

# set more random effect and error parameters

tau_1 <- 25 # by-subject random slope sd for random slopes

rho <- .3 # correlation between intercept and slope

sigma <- 30 # residual (error) sd

#But note that we are sampling two random effects for each subject s, a random intercept T0s and a random slope T1s. It is possible for these values to be positively or negatively correlated, in which case we should not sample them independently. For instance, perhaps people who are faster than average overall (negative random intercept) also show a smaller than average effect of the in-group/out-group manipulation (negative random slope) because they allocate less attention to the task. We can capture this by allowing for a small positive correlation between the two factors, rho, which we assign to be .2.

# overall model: RTsi=β0+T0s+O0i+(β1+T1s)Xi+esi. The response time for subject s on item i, RTsi, is decomposed into a population grand mean, β0; a by-subject random intercept, T0s; a by-item random interce
# set number of subjects and items

n_subj <- 16 # number of subjects
# pt, O0i; a fixed slope, β1; a by-subject random slope, T1s; and a trial-level residual, esi. Our data-generating process is fully determined by seven population parameters, all denoted by Greek letters: β0, β1, τ0, τ1, ρ, ω0, and σ (see Table 2). In the next section, we apply this data-generating process to simulate the sampling of subjects, items, and trials (encounters).


n_seated <- 6 # number of seated observations

n_standing <- 6 # number of standing observations

#We need to create a table listing each item i, which category it is in, and its random effect, O0i:

# simulate a sample of items

# total number of items = n_ingroup +n_outgroup

items <- data.frame(
  item_id = seq_len(n_seated + n_standing),
  category = rep(c("seated", "standing"), c(n_seated, n_standing)),
  O_0i = rnorm(n = n_seated + n_seated, mean = 0, sd = omega_0)

)

# effect-code category. this encodes a predictor as to which category each sim belongs to. seated should be less pwr, standing higher

items$X_i <- recode(items$category, "seated" = -0.5, "standing" = +0.5)
#We will later multiply this effect-coded factor by the fixed effect of category (beta_1 = 50) to simulate data in which the powers differ by postrue

# simulate a sample of subjects

# calculate random intercept / random slope covariance

covar <- rho * tau_0 * tau_1

# put values into variance-covariance matrix

cov_mx <- matrix(c(tau_0^2, covar, covar, tau_1^2),
                 nrow = 2, byrow = TRUE)

# generate the by-subject random effects

subject_rfx <- MASS::mvrnorm(n = n_subj, 
    mu = c(T_0s = 0, T_1s = 0),
    Sigma = cov_mx)

# combine with subject IDs

subjects <- data.frame(subj_id = seq_len(n_subj),subject_rfx)


# cross subject and item IDs; add an error term

# nrow(.) is the number of rows in the table

trials <- crossing(subjects, items) %>%
  mutate(e_si = rnorm(nrow(.), mean = 0, sd = sigma)) %>%
  select(subj_id, item_id, category, X_i, everything())

# calculate the response variable

dat_sim <- trials %>% 
  mutate(pwr = beta_0 + T_0s + O_0i + (beta_1 + T_1s) * X_i + e_si) %>%
  select(subj_id, item_id, category, X_i, pwr)

```

## Data visualization & Figure 1
```{r}

# Visualize the data
p1 <- ggplot(data = dat_sim, aes(x = pwr, fill = category)) + geom_density() +
  facet_wrap(~subj_id) + xlab('Power (W)') + ylab('Density') + ggtitle('LMEM Approach') #+ 
  #scale_fill_grey()
  
ggplot(data = dat_sim, aes(x = category, y = pwr, fill = category)) + geom_boxplot() +
  geom_point(position=position_jitterdodge())

dat_sim%>%
  group_by(subj_id, category)%>%
  summarize(
    meanPow = mean(pwr)
  ) %>%
  ggplot(aes(x = category, y = meanPow)) +   
  geom_point(aes(color=as.factor(subj_id))) + 
  geom_line(aes(group = subj_id, color = as.factor(subj_id) )) +
  theme_bw()
 

p2 <- dat_sim%>%
  group_by(subj_id, category)%>%
  summarize(
    meanPow = mean(pwr)
  ) %>%
  ggplot(aes(x = category, y = meanPow)) +   
  geom_point(aes(color=category)) +
  facet_wrap(~subj_id) + ylab('Mean Power (W)') +
  xlab('Posture') + ggtitle('Averaged Data Approach') #+
  #scale_color_grey() 

# Figure 1
p1 | p2

```

## Replicate measurements and Figure 2
```{r}
experiments <- 1e4
randDat <- rnorm(1000, mean = 800, sd = 50)
# function to calculate row varuances
rowVars <- function(x, na.rm=F) {
    # Vectorised version of variance filter
    rowSums((x - rowMeans(x, na.rm=na.rm))^2, na.rm=na.rm) / (ncol(x) - 1)
}

meanVec = numeric(length = 30)
sdVec = numeric(length = 30)

# Repeat from 1:30 grabbing values from 1 through 30 samples from the above distribution
for (noReps in 1:30){
  largeDat <- matrix(nrow = 1e4, ncol=noReps)
  for (i in 1:experiments){
    largeDat[i,1:noReps] <- sample(randDat, noReps, replace = TRUE)
  }
  rowMeans(largeDat)
  rowVars(largeDat)
  meanVec[noReps] <- mean( abs( (rep(mean(randDat),length(rowMeans)) - rowMeans(largeDat)) / sqrt(rowVars(largeDat)) ) )
  sdVec[noReps] <- sd( abs( (rep(mean(randDat),length(rowMeans)) - rowMeans(largeDat)) / sqrt(rowVars(largeDat)) ) )
}

replicateStudy <- data.frame(meanVec, sdVec, c(1:30))
names(replicateStudy)[3] <- 'NumReps'
ggplot(replicateStudy, aes(x=NumReps, y = meanVec)) + geom_point() +
  ylab('Sampling Error (|Z-Score|)') + xlab('Number of Replicates')
```

## loading unbalanced dat from python creation
```{r}
## to remove: subj 1 half of trials, subj
dat_mess <- read.csv('C:/Users/daniel.feeney/OneDrive - Boa Technology Inc/Desktop/Rethinking Neuro Data Manuscript/Code/Data/messDat.csv')
dat_unbal <- read.csv('C:/Users/daniel.feeney/OneDrive - Boa Technology Inc/Desktop/Rethinking Neuro Data Manuscript/Code/Data/unbalDat.csv')


```
 
## create a LMER model
```{r}
dat_sim$Subject <- dat_sim$subj_id

# fit a linear mixed-effects model to data

mod_sim <- lmer(pwr ~ category + (1 + category | subj_id), 
                data = dat_sim)
# derive output from lmer
a <- summary(mod_sim, corr = FALSE)
r_sq <- r.squaredGLMM(mod_sim)


nullMod <- lmer(pwr ~ (category|subj_id), 
                data = dat_sim)

aovRes <- anova(mod_sim, nullMod)
aovRes$`Pr(>Chisq)`[2]

```



## better RM model taken into account
```{r}
avgDat <- dat_sim %>%
  group_by(subj_id, category) %>%
  summarize( meanPwr = mean(pwr))

t.test(meanPwr ~ category, data = avgDat, paired = TRUE)
```

## Visualizations of no pooling, complete pooling, and partial pooling
```{r}

ggplot(data = dat_sim, mapping=aes(x=category, y=pwr, color = category))+
  geom_point() + facet_wrap(~subj_id)
  
# fit a no pooling model
no_pooling_mod <- lmList(pwr ~ category | subj_id, dat_sim)

df_no_pooling <- lmList(pwr ~ category | subj_id, dat_sim) %>% 
  coef() %>% 
  # Subject IDs are stored as row-names. Make them an explicit column
  rownames_to_column("Subject") %>% 
  rename(Intercept = `(Intercept)`, Slope_Cond = categorystanding) %>% 
  add_column(Model = "No pooling")
df_no_pooling$Subject <- as.integer(df_no_pooling$Subject)

#fit a complete pooling mode
m_pooled <- lm(pwr ~ category, dat_sim) 

# Repeat the intercept and slope terms for each participant
df_pooled <- tibble(
  Model = "Complete pooling",
  Subject = unique(dat_sim$subj_id),
  Intercept = coef(m_pooled)[1], 
  Slope_Cond = coef(m_pooled)[2]
)

head(df_pooled)
```

## compare models. No pooling is RM ANOVA, complete pooling is LM
## 
```{r}
# https://www.linguisticsociety.org/sites/default/files/e-learning/class_3_slides.pdf
# Join the raw data so we can use plot the points and the lines.
df_models <- bind_rows(df_pooled, df_no_pooling) %>% 
  left_join(dat_sim, by = "Subject")

p_model_comparison <- ggplot(df_models) + 
  aes(x = category, y = pwr) + ylab('Power (W)') +
  # Set the color mapping in this layer so the points don't get a color
  geom_abline(
    aes(intercept = Intercept, slope = Slope_Cond, color = Model, linetype = Model),
    size = .75
  ) + 
  geom_point() +
  facet_wrap("Subject") +
  theme(legend.position = "top", legend.justification = "left") +
  scale_color_grey()

p_model_comparison 

```

```{r}
#mod_sim
df_partial_pooling <- coef(mod_sim)[["subj_id"]] %>% 
  rownames_to_column("Subject") %>% 
  as_tibble() %>% 
  rename(Intercept = `(Intercept)`, Slope_Cond = categorystanding) %>% 
  add_column(Model = "Partial pooling")

df_partial_pooling$Subject <- as.integer(df_partial_pooling$Subject)
df_models <- bind_rows(df_pooled, df_no_pooling, df_partial_pooling) %>% 
  left_join(dat_sim, by = "Subject")

# Replace the data-set of the last plot
part1 <- p_model_comparison %+% df_models
```


## Figure 3 and visualizing all 3 models to show shrinkage
```{r}
df_fixef <- tibble(
  Model = "Partial pooling (average)",
  Intercept = fixef(mod_sim)[1],
  Slope_Cond = fixef(mod_sim)[2]
)

df_shrink <- df_pooled %>% 
  distinct(Model, Intercept, Slope_Cond) %>% 
  bind_rows(df_fixef)
df_shrink

df_pulled <- bind_rows(df_no_pooling, df_partial_pooling)

part2 <- ggplot(df_pulled) + 
  aes(x = Intercept, y = Slope_Cond, color = Model, shape = Model) + 
  geom_point(size = 2) + 
  geom_point(
    data = df_shrink, 
    size = 5,
    show.legend = FALSE
  ) + 
  # Draw an arrow connecting the observations between models
  geom_path(
    aes(group = Subject, color = NULL), 
    arrow = arrow(length = unit(.02, "npc")),
    show.legend = FALSE
  )  + 
  #ggtitle("Pooling of regression parameters") + 
  xlab("Intercept estimate") + 
  ylab("Slope estimate") +
  scale_color_grey() +
  theme(legend.position="top")

part1 | part2
```


## Comparing effect size estimats 
```{r}

print("Estimated Effect Sizes")

est_slope = df_pooled$Slope_Cond

df_pooled$Subject

ggplot(df_pooled) + 
  aes(x=Intercept, y=Slope_Cond) + 
  geom_point()

print(paste("Pooled model = ", mean(est_slope) / sd(est_slope))) #inf because sd of a single number is 0 

est_slope_noPool = df_no_pooling$Slope_Cond

df_no_pooling$Subject

ggplot(df_no_pooling) + 
  aes(x=Intercept, y=Slope_Cond, label = Subject) + 
  geom_point()+ geom_text(hjust = 1, vjust = 1)

print(paste("No-pooling model = ", mean(est_slope_noPool) / sd(est_slope_noPool)))

est_slope_partialPool = df_partial_pooling$Slope_Cond

df_partial_pooling$Subject

ggplot(df_partial_pooling) + 
  aes(x=Intercept, y=Slope_Cond, label = Subject) + 
  geom_point() + geom_text(hjust = 1, vjust = 1)

print(paste("Partial-pooling model = ", mean(est_slope_partialPool) / sd(est_slope_partialPool)))

a <- avgDat %>% filter(category=="seated")
a <- a$meanPwr

b <- avgDat %>% filter(category=="standing")
b <- b$meanPwr

print(paste("Paired t-test = ", mean(b-a) / sd(b-a)))



```

## plotting in R
```{r}

df_all <- bind_rows(df_pooled, df_no_pooling, df_partial_pooling)

ggplot(df_all) + 
  aes(x = Subject, y = Slope_Cond, color=Model, label = Subject) + 
  geom_point(size = 2) + 
  geom_point(
    data = df_all, 
    size = 2,
    # Prevent size-5 point from showing in legend keys
    show.legend = FALSE
  ) +
  geom_text(vjust=1, hjust=1)+
  scale_x_continuous(breaks=seq(0,15,1)) + 
  theme(panel.grid.minor = element_blank()) + 
  facet_grid(cols = vars(Model))


```

## Comparing messy dataset & Figure 4
```{r}
library(ciTools)
pp_mod <- lmer(pwr ~ category + (category|subj_id), data = dat_mess)
summary(pp_mod)
ciMod <- add_ci(dat_mess, pp_mod, alpha = 0.5)
coefsPP <- coef(pp_mod)
data.frame(coefsPP$subj_id)

cp_mod <- lm(pwr ~ category, data = dat_mess)
summary(np_mod)

np_mod <- lmList(pwr ~ category | subj_id, dat_mess) # Will fail based on 
np_mod

# plotting Figure 4
dat_mess$category <- as.factor(dat_mess$category)
dat_mess$subj_id <- as.factor(dat_mess$subj_id) 


ciMod %>%
  arrange(pwr)%>%
  group_by(subj_id, category) %>%
  filter(category == 'standing')%>%
  summarize(
    meanP = mean(pwr),
    predP = mean(pred),
    u25 = mean(LCB0.25),
    u75 = mean(UCB0.75))%>%
  ggplot(aes(x = fct_reorder(subj_id, meanP), y = meanP, ymin=u25, ymax=u75)) + 
  geom_point() + geom_errorbar()
```



# Appendix 
## add another fixed effect to show the model's ability to generalize
```{r}

dat_sim$footwear <- rep(c(rep('shoe1',3), rep('shoe2',3)),32)

ggplot(dat_sim, aes(x = category, y = pwr, color = footwear, fill = footwear)) +
  geom_point() + geom_jitter(width = 0.1, height = 0.1) + facet_wrap(~Subject) +
  scale_color_manual(values=c("red","blue")) + ylab("Power (W)") + xlab("Posture")

twoMod <- lmer(pwr ~ category + footwear + (category|Subject), data = dat_sim)
summary(twoMod)
```

## conduct F-test between anova models. In this case, there is no estimated main effect of footwear but there is a main effect of category (seated or standing)
```{r}

fwMod <- lmer(pwr ~ category + footwear + (category|Subject), data = dat_sim)
baseMod <- lmer(pwr ~ category + (category|Subject), data = dat_sim)
anova(fwMod, baseMod)

catMod <- lmer(pwr ~ category + footwear + (category|Subject), data = dat_sim)
baseMod <- lmer(pwr ~ footwear + (category|Subject), data = dat_sim)
anova(fwMod, baseMod)
```


## add interaction effects and plot them
```{r}
library(sjPlot)
intMod <- lmer(pwr ~ category * footwear + (category|Subject), data = dat_sim)
plot_model(intMod) 

plot_model(intMod, type = 'pred', terms = c("category", "footwear"))

summary(intMod)

```

## Compare to only a linear model (worst choice)     
```{r}
# complete pooling
wrongMod <- lm(pwr ~ category, data = dat_sim)
wrongSum <- summary(wrongMod)
wrongSum$coefficients[8]
wrongSum

# repeated measures error model is better #
aovMod <- aov( pwr ~ category + Error(factor(subj_id)), data = dat_sim )#
RMsum <- summary(aovMod)
RMsum$`Error: Within`
summary(aovMod)
```
 


